Community Detection for the\\ Generalized Random Dot Product Graph

### 2021-06-14 proposal/prospectus.Rmd
minor fix
updated prospectus

Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision. One common goal of analyzing graph data is community detection or graph clustering, in which the graph is partitioned into disconnected subgraphs in an unsupervised yet meaningful manner (e.g., by optimizing an objective function or recovering unobserved labels). Because traditional clustering techniques were developed for data that can be represented as vectors, they cannot be applied directly to graphs. Common examples of such clustering techniques include $K$-means clustering \cite{1056489} and Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}. In addition, these techniques assume that the vectors are sampled from a specific type of distribution (whether implicitly in the case of $K$-means clustering or explicitly in the case of GMM). In this research, we investigate the use of a family of spectral decomposition based approaches for community detection in block models (random graph models with inherent community structure). First, we demonstrate that under the Generalized Random Dot Product Graph \cite{rubindelanchy2017statistical} framework, all graphs generated by Block Models (generative graph models with inherent community structure) can be represented as collections of vectors in Euclidean space via the Adjacency Spectral Embedding \cite{athreya2017statistical, lyzinski2014}. Then, noting the particular structure or configuration that the vectors take, we select appropriate clustering techniques to apply on the vector representation of the graph. Finally we derive the asymptotic properties of these methods to show that our methods are consistent or achieve desirable properties with high probability. To illustrate this approach, we primarily focus on the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}.

@@

The SBM, DCBM, and PABM are all special cases of this generative model. Based on the structure of the $\gamma_k$ functions that correspond to each of the SBM, DCBM, and PABM, there are clear choices for clustering algorithms on the embeddings of each of these Block Models. Extensions of these results may consider which clustering algorithms result in consistent estimators given restrictions on each of the $\gamma_k$, and how to approach community detection when $\gamma_k$ are unknown altogether.

@@

1. Literature review and complete proofs of main theorems: December 2022

>>>>

Graph and network data, in which samples are represented not as a collection of feature vectors but as relationships between pairs of observations, are increasingly widespread in various fields ranging from sociology to computer vision. Analyzing such data, however, can be a challenge because traditional statistical and machine learning techniques are often incompatible with graph data. For example, if we observe data sampled from a mixture of Gaussians, we can iteratively update the means and covariances by computing summary statistics for each mixture component and update community labels via Mahalanobis distance using the sample means and covariances \cite{doi:10.1198/016214502760047131}. A type of model that is analogous to mixture models for graph data is the Block Model, which is a family of generative models that have community structure. For such models, it is often impossible to compute analogous summary statistics (e.g., sample mean) for community detection or parameter estimation. In this research, we investigate the use of a family of spectral decomposition and embedding based approaches for community detection in block models (random graph models with inherent community structure). First, we demonstrate that under the Generalized Random Dot Product Graph \cite{rubindelanchy2017statistical} framework, all graphs generated by Block Models (generative graph models with inherent community structure) can be represented as collections of vectors in Euclidean space via the Adjacency Spectral Embedding \cite{athreya2017statistical, lyzinski2014}. Then, noting the particular structure or configuration that the vectors take, we select appropriate clustering and parameter estimation techniques to apply on the vector representation of the graph. Finally we derive the asymptotic properties of these methods to show that our methods are consistent or achieve desirable properties with high probability. To illustrate this approach, we primarily focus on the Popularity Adjusted Block Model \cite{307cbeb9b1be48299388437423d94bf1}.

@@

The SBM, DCBM, and PABM are all special cases of this generative model. Based on the structure of the $\gamma_k$ functions that correspond to each of the SBM, DCBM, and PABM, there are natural choices for clustering algorithms on the embeddings of each of these Block Models, followed by straightforward parameter estimation techniques. Extensions of these results may consider which clustering algorithms result in consistent estimators given restrictions on each of the $\gamma_k$, and how to approach community detection when $\gamma_k$ are unknown altogether.

@@

1. Literature review and proofs of main theorems: December 2022

### 2021-06-14 proposal/prospectus.tex
minor fix
updated prospectus

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdfauthor={John Koo},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

@@

Graph and network data, in which samples are represented not as a
collection of feature vectors but as relationships between pairs of
observations, are increasingly widespread in various fields ranging from
sociology to computer vision. One common goal of analyzing graph data is
community detection or graph clustering, in which the graph is
partitioned into disconnected subgraphs in an unsupervised yet
meaningful manner (e.g., by optimizing an objective function or
recovering unobserved labels). Because traditional clustering techniques
were developed for data that can be represented as vectors, they cannot
be applied directly to graphs. Common examples of such clustering
techniques include \(K\)-means clustering \cite{1056489} and Gaussian
Mixture Models \cite{doi:10.1198/016214502760047131}. In addition, these
techniques assume that the vectors are sampled from a specific type of
distribution (whether implicitly in the case of \(K\)-means clustering
or explicitly in the case of GMM). In this research, we investigate the
use of a family of spectral decomposition based approaches for community
detection in block models (random graph models with inherent community
structure). First, we demonstrate that under the Generalized Random Dot
Product Graph \cite{rubindelanchy2017statistical} framework, all graphs
generated by Block Models (generative graph models with inherent
community structure) can be represented as collections of vectors in
Euclidean space via the Adjacency Spectral Embedding
\cite{athreya2017statistical, lyzinski2014}. Then, noting the particular
structure or configuration that the vectors take, we select appropriate
clustering techniques to apply on the vector representation of the
graph. Finally we derive the asymptotic properties of these methods to
show that our methods are consistent or achieve desirable properties
with high probability. To illustrate this approach, we primarily focus
on the Popularity Adjusted Block Model
\cite{307cbeb9b1be48299388437423d94bf1}.

@@

The SBM, DCBM, and PABM are all special cases of this generative model.
Based on the structure of the \(\gamma_k\) functions that correspond to
each of the SBM, DCBM, and PABM, there are clear choices for clustering
algorithms on the embeddings of each of these Block Models. Extensions
of these results may consider which clustering algorithms result in
consistent estimators given restrictions on each of the \(\gamma_k\),
and how to approach community detection when \(\gamma_k\) are unknown
altogether.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Literature review and complete proofs of main theorems: December 2022
\item
  Simulations, real data analyses, R package: March 2022
\item
  Dissertation Completion: April 2022
\end{enumerate}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdfauthor={John Koo},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}

@@

Graph and network data, in which samples are represented not as a
collection of feature vectors but as relationships between pairs of
observations, are increasingly widespread in various fields ranging from
sociology to computer vision. Analyzing such data, however, can be a
challenge because traditional statistical and machine learning
techniques are often incompatible with graph data. For example, if we
observe data sampled from a mixture of Gaussians, we can iteratively
update the means and covariances by computing summary statistics for
each mixture component and update community labels via Mahalanobis
distance using the sample means and covariances
\cite{doi:10.1198/016214502760047131}. A type of model that is analogous
to mixture models for graph data is the Block Model, which is a family
of generative models that have community structure. For such models, it
is often impossible to compute analogous summary statistics (e.g.,
sample mean) for community detection or parameter estimation. In this
research, we investigate the use of a family of spectral decomposition
and embedding based approaches for community detection in block models
(random graph models with inherent community structure). First, we
demonstrate that under the Generalized Random Dot Product Graph
\cite{rubindelanchy2017statistical} framework, all graphs generated by
Block Models (generative graph models with inherent community structure)
can be represented as collections of vectors in Euclidean space via the
Adjacency Spectral Embedding
\cite{athreya2017statistical, lyzinski2014}. Then, noting the particular
structure or configuration that the vectors take, we select appropriate
clustering and parameter estimation techniques to apply on the vector
representation of the graph. Finally we derive the asymptotic properties
of these methods to show that our methods are consistent or achieve
desirable properties with high probability. To illustrate this approach,
we primarily focus on the Popularity Adjusted Block Model
\cite{307cbeb9b1be48299388437423d94bf1}.

@@

The SBM, DCBM, and PABM are all special cases of this generative model.
Based on the structure of the \(\gamma_k\) functions that correspond to
each of the SBM, DCBM, and PABM, there are natural choices for
clustering algorithms on the embeddings of each of these Block Models,
followed by straightforward parameter estimation techniques. Extensions
of these results may consider which clustering algorithms result in
consistent estimators given restrictions on each of the \(\gamma_k\),
and how to approach community detection when \(\gamma_k\) are unknown
altogether.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Literature review and proofs of main theorems: December 2022
\item
  Simulations, real data analyses, R package: March 2022
\item
  Dissertation Completion: April 2022
\end{enumerate}
